A large and variable leading tail of Helium in a sub-Saturn undergoing runaway inflation

### 2023-06-05 paper/paper1/ms.tex
Fully remove the how we constructed OD contours paragraph
Add a paragraph on variability

\begin{figure}
    \includegraphics[width=0.8\linewidth]{figures/homogeneous.png}
    \caption{Isotropic mass loss}
    \label{fig:IsotropicMdot}
\end{figure}

\begin{figure}
    \includegraphics[width=0.8\linewidth]{figures/asym.png}
    \caption{Asymmetric dayside mass loss}
    \label{fig:DaysideMdot}
\end{figure}

@@

\subsection{3D Hydrodynamic simulations}\label{sec3Dsimulations}
\authorcomment1{Insert Morgan Macleod text here.}
\textcolor{lightgray}{Habitasse platea dictumst vestibulum rhoncus est pellentesque elit ullamcorper dignissim. Odio ut sem nulla pharetra. Odio ut sem nulla pharetra diam sit amet nisl suscipit. Accumsan tortor posuere ac ut. Ornare arcu dui vivamus arcu felis bibendum ut tristique. Nunc faucibus a pellentesque sit amet porttitor eget. Mauris rhoncus aenean vel elit. Sed vulputate mi sit amet mauris commodo. Quis lectus nulla at volutpat diam ut venenatis. Sed id semper risus in hendrerit gravida rutrum quisque non. Est placerat in egestas erat imperdiet sed. Viverra adipiscing at in tellus integer feugiat scelerisque varius. Quam viverra orci sagittis eu volutpat odio facilisis mauris sit. Morbi tristique senectus et netus et malesuada. Duis tristique sollicitudin nibh sit amet. Libero nunc consequat interdum varius sit amet mattis vulputate. Velit ut tortor pretium viverra suspendisse potenti nullam ac. Viverra accumsan in nisl nisi scelerisque eu ultrices vitae. Quis commodo odio aenean sed adipiscing diam donec.}

The overall observation of a leading tail comports with either of these two physical mechanisms, and we suspect both are at play in \hatpb.  Both cases appear needed to create the conditions of high enough column density to populate the \ion{He}{1} metastable triplet to detectable levels while imbuing the velocity substructure that we see.  Here we interpret the bulk velocity substructure in Figure \ref{fig:centroids} under a combination of the two.  We attribute the pre-transit arc to stellar wind acceleration.  The stellar wind is initially too weak to plow the dense escaping gas until a separation of 60-130 planetary radii ($+0.15$ orbital phase), the inflection point in an enormous bow shock.  Gas at these separations has had enough time to diffuse, both out of the orbital plane and to slightly larger radii shells, in turn lowering the column density along the line-of-sight.  The lower column density provides both fewer Helium atoms to participate in absorption, and less overall NUV shielding, leading to greater fractional ionization and further depopulation of the \ion{He}{1} metastable state.

@@

@@

\subsection{XUV Irradiation-driven runaway}
XUV photoevaporation heats the upper layers in the atmosphere, driving a hydrodynamic wind \citep{2009ApJ...693...23M}.  Photoevaporation stands out as offering a natural cause of the observed leading tail attributable to dayside/nightside differences in mass loss: the greatest supersonic motions arise from the sub-solar point on the planetary dayside.  The effect may be boosted if dayside/nightside energy transport proceeds inefficiently \citep{2009ApJ...693...23M}.  Such a scenario is illustrated in Figure \ref{fig:KeplerianShear}.

@@

\subsection{Ohmic Dissipation-driven runaway}
The Ohmic dissipation mechanism made two key observable predictions.  Heating efficiency should initially increase as a function of equilibrium temperature, then degrade as irradiation exceeds equilibrium temperatures of about 2000 K.  Second, hot Saturns ($\lesssim0.5 M_\mathrm{Jup}$) should undergo runaway evaporation, whereas hot Jupiters ($\gtrsim 1 M_\mathrm{Jup}$) should reach stable equilibrium---albeit inflated---radii after Gyr timescales.  Both of these outcomes predated data that could validate them, and these outcomes differ from the behavior of other heating mechanisms, such as photoevaporation or tides.

Ohmic dissipation stands out as a physical mechanism that could be responsible for both dramatic atmospheric escape and its significant radius inflation.  \citet{2011ApJ...738....1B} showed that Ohmic heating acts to inflate hot Jupiters, with planets $\lesssim 0.5 M_\mathrm{Jup}$ overflowing their Roche lobes and leading to evaporation on Gyr timescales.  The Ohmic heating scenario requires only a modest planetary magnetic field ($\gtrsim$1 G) and an equilibrium temperature great enough to thermally ionize some modest fraction of neutral metals, such as the low ionization species of \ion{Na}{1} and \ion{K}{1}.  The ``sweet spot'' for this phenomenon appears to prefer equilibrium temperatures in the range of $1500<T_\mathrm{eq}<2000$ K \citep{2011ApJ...738....1B,2012ApJ...745..138M,2016ApJ...819..116G,2018AJ....155..214T,2022A&A...658L...7K}, where the conductivity is strong enough to cause an effective drag without being so strong that magnetic braking slows the planetary wind.  HAT-P-67 b's $\sim$2000 K sits to the higher end, but still in a region of high Ohmic dissipation heating efficiency.  A proposed order-of-magnitude scaling law predicts an inflation timescale, Eq. 20 in \citet{2011ApJ...738....1B}:

@@

When applied to an ensemble of planet systems, both the XUV irradiation and Ohmic dissipation mechanisms expect a void in the planet mass-radius plane. However, the theories make different quantitative predictions in the placement of the dividing lines between stable and unstable populations.  We illustrate these differences in Figure \ref{fig:tf2018_theory}, which shades the mass-radius plane with predictions for the inflation timescale for planets with HAT-P-67b-like conditions.  The XUV irradiation timescale better matches the distribution of planets, with the 0.1 $g\;cm^{-3}$ density contour setting a conspicuous dividing line in density.  The Ohmic dissipation shading predicts short inflation timescales extending into a region of Jupiter-mass planets with radii commonly observed.  The better match of observed exoplanet demographics to the XUV irradiation shading disfavors Ohmic dissipation.

Under either scenario, HAT-P-67b resides in an extremely short-lifetime region.  HAT-P-32b offers an interesting test case since it has recently been shown to exhibit a large helium excess \authorcomment1{cite ZJ}.  It resides just slightly denser than the 0.1 $g\;cm^{-3}$ dividing line, with XUV predicting over 10$\times$ longer inflationary timescale for HAT-P-32b than for HAT-P-67b, while Ohmic dissipation expects merely a factor of 3 difference.  HAT-P-32b exhibits a more symmetric leading and trailing tail, whereas HAT-P-67b stands out as exhibiting evidence for preferential mass loss on the highly irradiated planet dayside.  These differences may make this pair an especially valuable laboratory for discriminating among theories of atmospheric escape.

@@

%We constructed the $\tau_\mathrm{infl}$ contours in the following way.  We censored the $\tau_\mathrm{infl}$ to regions in the mass-radius plane where numerical simulations of Ohmic dissipation resulted in runaway inflation \citet{2011ApJ...738....1B}.  We interpolate the $T_\mathrm{eq}$ trendlines from Fig. \ref{fig:ThornFortUpdated} to compute $\tau_\mathrm{infl}$ from Equation \ref{eqInflate}, including the best fit Gaussian prescription for $\hat{\epsilon}_G$ from Eq. 34 of \citet{2018AJ....155..214T}.  This trend should exhibit significant uncertainty owing to the order-of-magnitude nature of the scaling relation, source-to-source scatter in $\hat{\epsilon}_G$, and to what extent $\hat{\epsilon}_G$ applies for planets less than $0.5 M_\mathrm{Jup}$, which were excluded in the derivation of $\hat{\epsilon}_G$ owing to their scarcity.  Nevertheless, this contour appears to match the bulk trends: planets tend to reside in either the non-inflationary, stable locus or the inflationary-yet-stable locus.  A relatively few $\sim$dozens of planets approach the ``cosmic shoreline'' of inflationary lifetimes in the tens of Myr. The detection of atmospheric escape along this frontier (green circles in Figure \ref{fig:tf2018_theory}) supports this interpretation.  More observations and theoretical modeling of atmospheric escape along this ridge could help better understand its realized versus expected shape.

@@

We measured a large extent of Helium escape along the arc of the orbit, but we necessarily can place only coarse constraints on the vertical extent---in the direction perpendicular to the orbital path, $\pm z$.  We estimate the detectable vertical extent must be at least a few stellar radii, such that we still could have detected Helium escape in a hypothetical HAT-P-67b-like system even if it were non-transiting, in a ``near miss'' configuration.  We propose a new category of semi-transiting planet, dubbed ``exospheric grazers'', in which the planet does not produce a detectable white-light transit depth, but \emph{does} produce measurable \emph{line}-based exospheric absorption.  This category appears to have been neglected due to the assumption that exospheres were only easily detectable at several planetary radii, and therefore fractions of a stellar radius.  While large tails have been seen previously in Ly$\alpha$, the scarcity of UV resources prohibited searches of this kind.  The discovery of large Helium tails in HAT-P-32b (Zhang et al. 2023) and now HAT-P-67b suggest that the identification of these exospheric grazers may be achievable with current instrumentation and possibly existing archival data from near-IR RV-planet searches.  Non-transiting, strongly irradiated planets with well-constrained orbits may make ideal targets for detection of this phenomenon.

@@

\begin{figure}[b]
    \includegraphics[width=0.5\linewidth]{figures/Ca8662_phase_2D_diagram_resid.png}
    \caption{Non-detection of variability in the Calcium IR Triplet line at 8862 $\AA$.  No other HPF lines appear to show any variability.}
    \label{fig:CaPhaseScan}
\end{figure}

>>>>

\begin{figure}
    \includegraphics[width=0.8\linewidth]{figures/homogeneous.png}
    \caption{3D hydrodynamic simulations of HAT-P-67b with isotropic mass loss.  The upper image shows density along a slice in the orbital plane.  The bottom two panels show trens in orbital phase, with (top) column density and (bottom) line-of-sight velocity.}
    \label{fig:IsotropicMdot}
\end{figure}

\begin{figure}
    \includegraphics[width=0.8\linewidth]{figures/asym.png}
    \caption{Same as \ref{fig:IsotropicMdot}, but with mass loss concentrated to the planet dayside.  The substellar point exhibits the greatest mass loss, reducing to near zero at the planet terminator.  A predominantly leading tail is produced.}
    \label{fig:DaysideMdot}
\end{figure}

@@

\subsection{3D Hydrodynamic simulations}\label{sec3Dsimulations}
We explore the observational impact of dayside mass loss with 3D hydodynamic models.  The mechanics follow previous such simulations on WASP 107~b \citep{2022ApJ...926..226M} and HAT-P-32~b (Zhang et al. 2023).  We modified the simulations to match the system properties of HAT-P-67~b.  We ran two simulations: one in which the planetary boundary conditions are identical everywhere on the sphere of the planet (the homogeneous or isotropic case). In the
other (asymmetric or anisotropic case) we let the temperature correspond to the incident irradiation through a $\cos(\theta)$ term where $\theta$ is the substellar angle.  Figures \ref{fig:IsotropicMdot} and \ref{fig:DaysideMdot} compare these two scenarios.  The dayside heating case clearly produces a large leading tail comparable to the scale observed in the Helium excess absorption.
\authorcomment1{Insert Morgan Macleod text here.}

Orbtial shear, stellar wind confinement, and preferential mass loss may all consipire to create the conditions of high enough column density to populate the \ion{He}{1} metastable triplet to detectable levels while imbuing the velocity substructure that we see.  Here we interpret the bulk velocity substructure in Figure \ref{fig:centroids} under these gas dynamics mechanisms.  We attribute the pre-transit arc to stellar wind acceleration.  The stellar wind is initially too weak to plow the dense escaping gas until a separation of 60-130 planetary radii ($+0.15$ orbital phase), the inflection point in an enormous bow shock.  Gas at these separations has had enough time to diffuse, both out of the orbital plane and to slightly larger radii shells, in turn lowering the column density along the line-of-sight.  The lower column density provides both fewer Helium atoms to participate in absorption, and less overall NUV shielding, leading to greater fractional ionization and further depopulation of the \ion{He}{1} metastable state.

@@

\subsection{Variable planetary wind}
Figure \ref{fig:centroids} and \ref{fig:HPFperCampaign} show differences in the Helium line profiles over months-long and years-long timescales.  The line profiles are qualitatively similar, but show slightly different tendencies towards redshifting.  For example, the centroid of the Helium line during the April 2022 transit appears slightly blueshifted relative to the stellar rest frame.  In comparison, the May 2020 transit probes the same planetary phases, yet resides slightly redshifted at transit midcenter.  We interpret this line profile variability as indicative of genuine planetary wind variability, which can arise from the interplay of stellar and planetary winds \citep{2009ApJ...693...23M} and weather-like feedbacks in the planetary upper atmosphere.

@@

\subsection{XUV Irradiation-driven mass loss}
XUV photoevaporation heats the upper layers in the atmosphere, driving a hydrodynamic wind \citep{2009ApJ...693...23M}.  Photoevaporation stands out as offering a natural cause of the observed leading tail attributable to dayside/nightside differences in mass loss: the greatest supersonic motions arise from the sub-solar point on the planetary dayside.  The effect may be boosted if dayside/nightside energy transport proceeds inefficiently \citep{2009ApJ...693...23M}.  Such a scenario is illustrated in Figure \ref{fig:KeplerianShear}.

@@

\subsection{Ohmic Dissipation-driven mass loss}
The Ohmic dissipation mechanism made two key observable predictions.  Heating efficiency should initially increase as a function of equilibrium temperature, then degrade as irradiation exceeds equilibrium temperatures of about 2000 K.  Second, hot Saturns ($\lesssim0.5 M_\mathrm{Jup}$) should undergo runaway evaporation, whereas hot Jupiters ($\gtrsim 1 M_\mathrm{Jup}$) should reach stable equilibrium---albeit inflated---radii after Gyr timescales.  Both of these outcomes predated data that could validate them, and these outcomes differ from the behavior of other heating mechanisms, such as photoevaporation or tides.

Ohmic dissipation may therefore be responsible for both dramatic atmospheric escape and significant radius inflation.  \citet{2011ApJ...738....1B} showed that Ohmic heating acts to inflate hot Jupiters, with planets $\lesssim 0.5 M_\mathrm{Jup}$ overflowing their Roche lobes and leading to evaporation on Gyr timescales.  The Ohmic heating scenario requires only a modest planetary magnetic field ($\gtrsim$1 G) and an equilibrium temperature great enough to thermally ionize some modest fraction of neutral metals, such as the low ionization species of \ion{Na}{1} and \ion{K}{1}.  The ``sweet spot'' for this phenomenon appears to prefer equilibrium temperatures in the range of $1500<T_\mathrm{eq}<2000$ K \citep{2011ApJ...738....1B,2012ApJ...745..138M,2016ApJ...819..116G,2018AJ....155..214T,2022A&A...658L...7K}, where the conductivity is strong enough to cause an effective drag without being so strong that magnetic braking slows the planetary wind.  HAT-P-67 b's $\sim$2000 K sits to the higher end, but still in a region of high Ohmic dissipation heating efficiency.  A proposed order-of-magnitude scaling law predicts an inflation timescale, Eq. 20 in \citet{2011ApJ...738....1B}:

@@

When applied to an ensemble of planet systems, both the XUV irradiation and Ohmic dissipation mechanisms expect a void in the planet mass-radius plane. However, the theories make different quantitative predictions for the placement of the dividing lines between stable and unstable populations.  We illustrate these differences in Figure \ref{fig:tf2018_theory}, which shades the mass-radius plane with predictions for the inflation timescale under HAT-P-67b-like conditions.  The XUV irradiation timescale better matches the distribution of planets, with the 0.1 $g\;cm^{-3}$ density contour setting a conspicuous dividing line in density.  The Ohmic dissipation shading predicts short inflation timescales extending into a region of Jupiter-mass planets with radii commonly observed.  The better match of observed exoplanet demographics to the XUV irradiation shading disfavors Ohmic dissipation.

Under either scenario, HAT-P-67b resides in an extremely short-lifetime region.  Its nearest analog, HAT-P-32b, offers an interesting test case since it has recently been shown to exhibit a large helium excess \authorcomment1{cite ZJ}.  It resides just slightly denser than the 0.1 $g\;cm^{-3}$ dividing line, with XUV predicting over 10$\times$ longer inflationary timescale for HAT-P-32b than for HAT-P-67b, while Ohmic dissipation expects merely a factor of 3 difference.  HAT-P-32b exhibits a more symmetric leading and trailing tail, whereas HAT-P-67b stands out as exhibiting evidence for preferential mass loss on the highly irradiated planet dayside.  These differences may make this pair an especially valuable laboratory for developing theories of atmospheric escape.

@@

@@

We measured a large extent of Helium escape along the arc of the orbit, but we necessarily can place only coarse constraints on the vertical extent---in the direction perpendicular to the orbital path, $\pm z$.  We estimate the detectable vertical extent must be at least a few stellar radii, such that we still could have detected Helium escape in a hypothetical HAT-P-67b-like system even if it were non-transiting, in a ``near miss'' configuration.  We propose a new category of semi-transiting planet, dubbed ``exospheric grazers'', in which the planet does not produce a detectable white-light transit depth, but \emph{does} produce measurable \emph{line}-based exospheric absorption.  This category appears to have been neglected due to the assumption that exospheres were only easily detectable at several planetary radii.  While large tails have been seen previously in Ly$\alpha$, the scarcity of UV resources prohibited searches of this kind.  The discovery of large Helium tails in HAT-P-32b (Zhang et al. 2023) and now HAT-P-67b suggest that the identification of these exospheric grazers may be achievable with current instrumentation and possibly existing archival data from near-IR RV-planet searches.  Non-transiting, strongly irradiated planets with well-constrained orbits may make ideal targets for detection of this phenomenon.

@@

\begin{figure}[b]
    \includegraphics[width=0.5\linewidth]{figures/Ca8662_phase_2D_diagram_resid.png}
    \caption{Non-detection of variability in the Calcium IR Triplet line at 8862 \AA.  No other HPF lines appear to show any variability.}
    \label{fig:CaPhaseScan}
\end{figure}
